---
title: "R Coding Exercise"
author: Mohammed Zuber
date: January 21, 2025
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
---

Placeholder file for the future R coding exercise.


```{r, echo=FALSE, message=FALSE}
# loading the required R packages
library(here)
# the package dslabs will load the gapminder data
library(dslabs)
library(tidyverse) 
library(knitr)
```



## Loading and Checking Data

```{r}
# help() function pulls up the help page for the data to see what it contains.
help(gapminder) 
# str () give us the overview of the datasets
str(gapminder) 
# summary () gives summary of data
summary(gapminder) 
#class() function to check what type of object gapminder is
class(gapminder) # obtaining the type of object gapminder is
```

## Processing Data
```{r}
#filter african countries to make a new dataset called 'africadata'
africadata <- gapminder %>% filter(continent == "Africa") 
# getting an overview of the data structure  africadata
str(africadata) 
# summary of africadata
summary(africadata) 
```
```{r}
# creating two new objects for further analysis 
variableone <- africadata %>% select(infant_mortality, life_expectancy) 
# obtaining the information about the structure of variableone
str(variableone) 
# summarizing the details about variableone
summary(variableone) 
# creating another dataset with only two variables as listed
variabletwo <- africadata %>% select(population, life_expectancy) 
# obtaining the information about the structure of variabletwo
str(variabletwo )
# obtaining the summary about the dataset variabletwo
summary(variabletwo)  
```

## Plotting 
### Figure 1 Scatterplot of Life Expectancy vs Infant Mortality 
The scatterplot shows the relationship between life expectancy and infant mortality.The scattered plots shows that the low infant mortality tend to have higher life expectancy, indicating better healthcare, sanitation, and economic conditions. Whereas high infants mortality rates are associated with lower life expectancy, possibly due to poor healthcare access, malnutrition, and economic hardships.

```{r}
# Create scatterplot life expectancy as a function of infant mortality
ggplot(variableone, aes(x = log(infant_mortality), y = life_expectancy)) + 
  geom_point(color = "green", alpha = 0.7) +
labs(
    title = "Scatterplot of Life Expectancy vs Infant Mortality",
    x = "Infant Mortality",
    y = "Life Expectancy"
  ) +
  theme_minimal()

```
### Figure 2: Scatterplot of Life Expectancy vs Population Size

The scatterplot depicts the relationship between life expectancy and population size. There is some positive correlation between population size and life expectancy, but it is not strongly linear. Larger population might be associated with better healthcare infrastructure, economic growth, and development, leading to higher life expectancy.However, some small countries still exhibit high life expectancy, indicating that factors like healthcare access and living conditions play a crucial role beyond population size.
```{r}
# Create scatterplot 
ggplot(variabletwo, aes(x = log(population), y = life_expectancy)) + 
  geom_point(color = "purple", alpha = 0.7) +
labs(
    title = "Scatterplot of Life Expectancy vs Population Size",
    x = "Population Size (Log Scale)",
    y = "Life Expectancy"
  ) +
  theme_minimal()

```
### Table 1: Summary of missing values for infant mortality

```{r}
# Calculating the number of missing values for infant mortality by year
missing_data <- africadata %>% group_by(year) %>% summarize( missing_infant_mortality = sum(is.na(infant_mortality)), 
    total_observations = n(), #this will gives us the total observations
  )

# printing the result 
print(missing_data)

```
```{r}
# Let's create new dataset containing only the year 2000
data_new <- africadata %>% filter(year == 2000)
str(data_new)
summary(data_new)
```
### Figure 3: Infant mortality vs life expectancy scatterplot for 2000

```{r}
# Creating a scatterplot 
ggplot(data_new, aes(x = log(infant_mortality), y = life_expectancy)) + 
  geom_point(color = "brown", alpha = 0.7) + geom_smooth(method = "lm", color = "blue", se = FALSE) +
labs(
    title = "Scatterplot of Life Expectancy vs Infant Mortality for year 2000",
    x = "Infant Mortality",
    y = "Life Expectancy"
  ) +
  theme_minimal()

```
### Figure 4: Infany mortality vs population catterplot for year 2000

```{r}
# Creating scatterplot 
ggplot(data_new, aes(x = log(population), y = life_expectancy)) + 
  geom_point(color = "black", alpha = 0.7) + geom_smooth(method = "lm", color = "blue", se = FALSE) +
labs(
    title = "Scatterplot of Life Expectancy vs Population Size for the year 2000",
    x = "Population Size (Natural Log Scale)",
    y = "Life Expectancy"
  ) +
  theme_minimal()

```
## Fitting a linear model

In the first model, the p-value is less than 0.05. Therefore, this suggest that infant mortality is a strong predictor of life expectancy.the R-squared of 0.4701 means that 47% of the variation in life expectancy is explained by infant mortality.

In the second model, the p-value is greater than 0.05. Therefore, there is no strong statistical evidence to suggest that population size is a significant predictor of life expectancy.The R-squared of 0.01044 means that 1% of the variation in life expectancy is explained by population size, which is weak. 

```{r}
# Fitting the linear model
fit1 <- lm(life_expectancy ~ infant_mortality, data = data_new)
fit2 <- lm(life_expectancy ~ log(population), data = data_new)
# summary of the model
summary(fit1)
summary(fit2) 
```

