---
title: "Synthetic Data Exercise"
author: Mohammed Zuber
date: January 30, 2025
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
---

## Introduction
For this exercise, I simulated synthetic data for a cancer treatment cohort to study the associations between patient characteristics and treatment outcomes. The dataset includes 500 patients receiving one of four treatments: Chemotherapy, Immunotherapy, Targeted Therapy, or CAR-T Therapy.

The following variables are included:

PatientID – Unique identifier for each patient.
Age – Age at cancer diagnosis (years).
Sex – Male or Female.
FamilyHistory – Whether the patient has a family history of cancer (binary).
SmokingStatus – Whether the patient is a current smoker (binary).
BMI – Patient’s Body Mass Index (BMI).
OS_Months – Overall survival in months, influenced by patient characteristics and treatment type.
PFS_Months – Progression-free survival, derived as a proportion of OS.
CompleteResponse (CR) – Binary variable indicating if the patient achieved a complete response to treatment.
AdverseEvent (ADR) – Binary variable indicating if the patient experienced an adverse event.


```{r setup, include=FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(here)
# Set seed for reproducibility
set.seed(123)
```

## Generate Synthetic Data**

```{r data-generation}
# Number of patients
num_patients <- 500  

# Define treatment groups
treatment_groups <- c("Chemotherapy", "Immunotherapy", "Targeted Therapy", "CAR-T Therapy")

# Generate patient IDs
patient_ids <- 1:num_patients

# Assign patients randomly to a treatment group
treatment_assignments <- sample(treatment_groups, num_patients, replace = TRUE)

# Generate baseline characteristics
age <- rnorm(num_patients, mean = 60, sd = 12)  # Age at diagnosis (Normal distribution)
sex <- sample(c("Male", "Female"), num_patients, replace = TRUE)  # Gender
family_history <- rbinom(num_patients, 1, prob = 0.3)  # 30% with family history
smoking_status <- rbinom(num_patients, 1, prob = 0.25)  # 25% current smokers
bmi <- rnorm(num_patients, mean = 27, sd = 5)  # BMI (Normal distribution)

# Function to generate Overall Survival (OS) based on characteristics
generate_os <- function(age, smoking, family_history, treatment) {
  base_os <- 36 - (age * 0.3) - (smoking * 5) + (family_history * 2)
  treatment_effect <- ifelse(treatment == "Chemotherapy", -5, 
                      ifelse(treatment == "Immunotherapy", 2, 
                      ifelse(treatment == "Targeted Therapy", 4, 7)))
  return(base_os + treatment_effect + rnorm(1, mean = 0, sd = 5))
}

# Function to generate PFS as a fraction of OS
generate_pfs <- function(os) {
  return(runif(1, min = os * 0.4, max = os * 0.8))
}

# Function to generate Complete Response (CR)
generate_cr <- function(age, bmi, treatment) {
  prob_cr <- plogis(-0.05 * age + 0.03 * (30 - bmi) + 
                    ifelse(treatment == "Chemotherapy", -0.2, 
                    ifelse(treatment == "Immunotherapy", 0.1, 
                    ifelse(treatment == "Targeted Therapy", 0.2, 0.4))))
  return(rbinom(1, 1, prob = prob_cr))
}

# Function to generate Adverse Drug Reaction (ADR)
generate_adr <- function(age, smoking, bmi, treatment) {
  prob_adr <- plogis(0.03 * age + 0.1 * smoking + 0.02 * bmi + 
                     ifelse(treatment == "Chemotherapy", 0.3, 
                     ifelse(treatment == "Immunotherapy", 0.1, 
                     ifelse(treatment == "Targeted Therapy", 0.05, -0.1))))
  return(rbinom(1, 1, prob = prob_adr))
}

# Create dataset
cancer_data <- data.frame(
  PatientID = patient_ids,
  TreatmentGroup = treatment_assignments,
  Age = age,
  Sex = sex,
  FamilyHistory = family_history,
  SmokingStatus = smoking_status,
  BMI = bmi,
  OS_Months = mapply(generate_os, age, smoking_status, family_history, treatment_assignments),
  PFS_Months = NA,  # Placeholder
  CompleteResponse = mapply(generate_cr, age, bmi, treatment_assignments),
  AdverseEvent = mapply(generate_adr, age, smoking_status, bmi, treatment_assignments)
)

# Generate PFS values based on OS
cancer_data$PFS_Months <- mapply(generate_pfs, cancer_data$OS_Months)

# Ensure PFS is not greater than OS
cancer_data$PFS_Months <- pmin(cancer_data$PFS_Months, cancer_data$OS_Months)

# Print summary of dataset
summary(cancer_data)
```


## **Table: Summary Statistics by Treatment Group**
```{r summary-table}
cancer_data %>%
  group_by(TreatmentGroup) %>%
  summarise(
    Median_OS = median(OS_Months),
    Median_PFS = median(PFS_Months),
    CR_Rate = mean(CompleteResponse) * 100,
    ADR_Rate = mean(AdverseEvent) * 100
  )
```
## Exploratory Data Analysis**

```{r exploratory-plots}
# Boxplot: OS by Treatment Group
ggplot(cancer_data, aes(x = TreatmentGroup, y = OS_Months, fill = TreatmentGroup)) +
  geom_boxplot() +
  labs(title = "Overall Survival by Treatment Group",
       x = "Treatment Group",
       y = "Overall Survival (Months)") +
  theme_minimal()
# Proportion of Complete Response by Treatment Group
ggplot(cancer_data, aes(x = TreatmentGroup, fill = as.factor(CompleteResponse))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Complete Response by Treatment Group",
       x = "Treatment Group",
       y = "Proportion",
       fill = "Complete Response") +
  theme_minimal()
# Bar chart: Proportion of Adverse Events
ggplot(cancer_data, aes(x = TreatmentGroup, fill = as.factor(AdverseEvent))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Adverse Events by Treatment Group",
       x = "Treatment Group",
       y = "Proportion",
       fill = "Adverse Event") +
  theme_minimal()
```

## Fit Models to Recover Associations**

### **Linear Regression: Predicting OS**
```{r linear-model}
lm_os <- lm(OS_Months ~ Age + SmokingStatus + FamilyHistory + BMI + TreatmentGroup, data = cancer_data)
summary(lm_os)
```
Age and smoking significantly reduce overall survival (OS), while family history slightly improves it. BMI has no significant effect on OS. Chemotherapy shows the worst survival outcomes, followed by immunotherapy and targeted therapy. CAR-T therapy provides the best survival benefit, confirming its superior effectiveness.

### **Logistic Regression: Predicting Adverse Events**
```{r logistic-model}
glm_adr <- glm(AdverseEvent ~ Age + SmokingStatus + BMI + TreatmentGroup, 
               data = cancer_data, family = binomial)
summary(glm_adr)
```
Age is the only significant predictor, with older patients having a higher likelihood of experiencing ADRs. Smoking and BMI do not show a meaningful impact on ADR risk. While treatment effects are not strongly significant, immunotherapy shows a potential trend toward increased ADR risk compared to CAR-T therapy. Further research is needed to identify additional risk factors, such as comorbidities and genetic predisposition.

## Save Data**
```{r save-file}
# Save the simulated data to a CSV and Rds file
write.csv(cancer_data, here("synthetic-data-exercise","data","data.csv"), row.names = FALSE)
```