---
title: "CDC Data Exercise"
author: Mohammed Zuber
date: Febrary 4, 2025
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
---

# Introduction

## FAERS Data Description

The **FDA Adverse Event Reporting System (FAERS)** is a publicly available database maintained by the U.S. Food and Drug Administration (FDA). It contains adverse event reports, medication error reports, and product quality complaints submitted by healthcare professionals, manufacturers, and consumers. This data is used for post market surveillance of drug safety.

## Source of Data

The dataset is obtained from FAERS, which collects reports on suspected adverse drug reactions. The official FAERS database can be accessed at the FDA’s website:

[FDA FAERS Public Dashboard](https://www.fda.gov/drugs/questions-and-answers-fda-adverse-event-reporting-system-faers/fda-adverse-event-reporting-system-faers-public-dashboard)

## Contents of the Dataset

The uploaded FAERS dataset consists of **21 columns** that provide detailed information about adverse drug reactions, including:

- **Case ID**: Unique identifier for each reported case.
- **Suspect Product Names**: Drugs suspected to have caused an adverse event.
- **Suspect Product Active Ingredients**: Active pharmaceutical ingredients in the reported drugs.
- **Reason for Use**: The medical condition(s) for which the drug was prescribed.
- **Reactions**: Reported adverse events.
- **Serious**: Whether the adverse event is classified as serious.
- **Outcomes**: The final outcome of the reported event (e.g., hospitalization, death).
- **Sex**: Gender of the patient.
- **Event Date**: When the adverse event was reported.
- **Patient Age & Weight**: Demographics of the affected individual.
- **Reporter Type**: Whether the report was submitted by a healthcare professional or a consumer.
- **Concomitant Product Names**: Other drugs the patient was taking at the same time.
- **Country where Event Occurred**: Location of the reported adverse event.
- **Dates of FDA and Manufacturer Report Receipt**: Timeline of when the report was processed.

This dataset is valuable for analyzing post-market drug safety, identifying trends in adverse events, and informing regulatory actions.



```{r setup, include=FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(here)
library(kableExtra) # For enhanced table formatting
# Set seed for reproducibility
set.seed(123)
```

# Data Loading and Cleaning**
```{r}
# Define the file path
file_path <- here("cdcdata-exercise","data.csv")

# Function to load and clean data
load_and_clean_data <- function(file_path) {
  data <- read_csv(file_path, show_col_types = FALSE) %>%
    select("Suspect Product Active Ingredients", "Reason for Use", "Sex", 
           "Patient Age", "Patient Weight", "Outcomes", "Reactions") %>%
    
    # Keep only rows where 'Suspect Product Active Ingredients' contains 'Venetoclax'
    filter(str_detect(`Suspect Product Active Ingredients`, "Venetoclax")) %>%
    
    # Keep only rows where 'Reason for Use' is 'Chronic Lymphocytic Leukaemia'
    filter(`Reason for Use` == "Chronic Lymphocytic Leukaemia") %>%
    
    # Convert 'Patient Age' to numeric, removing 'YR' and filtering out 'Not Specified'
    mutate(`Patient Age` = na_if(`Patient Age`, "Not Specified")) %>%
    mutate(`Patient Age` = as.numeric(gsub(" YR", "", `Patient Age`))) %>%
    filter(!is.na(`Patient Age`)) %>%
    
    # Keep only single-word outcomes in 'Outcomes'
    filter(str_count(`Outcomes`, " ") == 0) %>%
    
    # Keep only single-word reactions in 'Reactions'
    mutate(`Reactions` = na_if(`Reactions`, "Not Specified")) %>%
    filter(str_count(`Reactions`, " ") == 0) %>%
    
    # Convert 'Patient Weight' to numeric, removing 'KG' and filtering out 'Not Specified'
    mutate(`Patient Weight` = na_if(`Patient Weight`, "Not Specified")) %>%
    mutate(`Patient Weight` = as.numeric(gsub(" KG", "", `Patient Weight`))) %>%
    filter(!is.na(`Patient Weight`))
  
  return(data)
}

# Load and clean the data
data <- load_and_clean_data(file_path)

# Check the structure and summary of the cleaned dataset

# Overview of the dataset structure
str(data) 

# Summary of the dataset
summary(data) 

# Check the class of the object
class(data)
```
# Descriptive analysis

```{r}
# Perform exploratory/descriptive analysis
summary_table <- data %>%
  summarise(
    count = n(),
    avg_age = mean(`Patient Age`, na.rm = TRUE),
    sd_age = sd(`Patient Age`, na.rm = TRUE),
    min_age = min(`Patient Age`, na.rm = TRUE),
    max_age = max(`Patient Age`, na.rm = TRUE),
    avg_weight = mean(`Patient Weight`, na.rm = TRUE),
    sd_weight = sd(`Patient Weight`, na.rm = TRUE),
    min_weight = min(`Patient Weight`, na.rm = TRUE),
    max_weight = max(`Patient Weight`, na.rm = TRUE)
  )

kable(summary_table) %>% 
  kable_styling(full_width = FALSE)
```



## Distribution of Patient Age
```{r}
ggplot(data, aes(x = `Patient Age`)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Patient Age", x = "Age", y = "Frequency") +
  theme_minimal()
```

## Distribution of Patient Weight
```{r}
ggplot(data, aes(x = `Patient Weight`)) +
  geom_histogram(binwidth = 5, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Patient Weight", x = "Weight (kg)", y = "Frequency") +
  theme_minimal()
```
## Percentage distribution of Sex
```{r}
sex_distribution <- data %>%
  count(Sex) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

print("Percentage Distribution of Sex:")
print(sex_distribution)
```

## Bar plot for Sex distribution
```{r}
ggplot(data, aes(x = Sex, fill = Sex)) +
  geom_bar() +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count") +
  theme_minimal()
```

## Frequency Table for Top 5 Adverse Reactions
```{r}
reaction_counts <- data %>%
  count(Reactions, name = "Frequency") %>%
  arrange(desc(Frequency)) %>%
  top_n(5, wt = Frequency)

print("Top 5 Most Frequent Adverse Reactions:")
print(reaction_counts)
```
## Percentage distribution of Outcomes
```{r}
outcome_distribution <- data %>%
  count(Outcomes) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

print("Percentage Distribution of Outcomes:")
print(outcome_distribution)
```
## Bar plot for Outcome frequencies
```{r}
ggplot(data, aes(x = Outcomes, fill = Outcomes)) +
  geom_bar() +
  labs(title = "Distribution of Outcomes", x = "Outcome", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# This portion is contributed by PRASANGA PAUDEL



Adding required packages
```{r}
library(purrr) #used during data generation
```

## Data Generation


### Setting the seeds and observation number
In this section we will create the baseline for generating our synthetic data. We will set the seed so that we can create reproducible data, and we will also set the number of observations to 135 as in the original dataset above.

```{r}
# Setting a seed for reproducibility of data
set.seed(100) # we will also set seed individually
# Definining the number of observations (patients) to generate
obs <- 135
```

### Creating an empty Dataset
We will then create an empty dataset with just variable names and their data structure. We will make sure that every variable has the same number of obserations i.e 135.

```{r}
# Create an empty data frame with placeholders for variables
synthetic <- data.frame(
  Weight = numeric(obs),#patient's weight
  Age = numeric(obs), #patient's age
  Sex = character(obs), #patient's sex
  Reaction = character(obs), #reported adverse events
  Outcome = character(obs), # final outcome of reported events
  activeIngredient = character(obs),
  Reason = character(obs) #reason for use
)
```

{{< pagebreak >}}

## Data Formation
Next, we will create the synthetic the data required for our excercise.
We will try to control how the data is formed while we generate these data. 
All the variables will be generated with as per the information obtained from the table and figures above.

We will also make sure that the variables contain mostly similar data when generated synthetically.


```{r}
# Variable 1: Weight (represents the weight of the patient)
synthetic$Weight <- round(rnorm(obs, mean = 77.58, sd = 17.27))
synthetic$Weight <- pmax(42.22, pmin(synthetic$Weight, 125.00))

# Variable 2: Age of the patient
set.seed(100) #setting an individual seed for age
adjusted_sd <- (88 - 77.6) / 1.75 # we observe that the tails dont have enough observations so we plan our data generation accordingly
synthetic$Age <- round(rnorm(obs, mean = 77.6, sd = adjusted_sd))
synthetic$Age <- pmax(30, pmin(synthetic$Age, 94)) # setting max and min

# Variable 3: Gender/Sex of the player
set.seed(123) #setting an individual seed for age
prob_female <- 0.2667  # 26.67%
prob_male <- 0.7259    # 72.59%
prob_not_specified <- 0.0074  # 0.74%

synthetic$Sex <- sample(
  c("Male", "Female", "Not Specified"),
  size = obs,
  replace = TRUE,
  prob = c(prob_male, prob_female, prob_not_specified) # setting probability same as original
)

# Define the reaction and frequency vectors
reaction <- c("Covid-19", "Pneumonia", "Death", "Infection", "Influenza",  
              "Abscess", "Appendicitis;Appendicectomy", "Arrhythmia", "Ascites", "Asthma",  
              "Bradycardia", "Bronchitis", "Bronchitis;Infection", "Cholecystitis;Gastritis;Cholecystectomy",  
              "Constipation", "Cough", "Covid-19;Sepsis", "Death;Thrombocytopenia",  
              "Dehydration", "Dehydration;Colitis;Diarrhoea", "Diarrhoea", "Diarrhoea;Fatigue",  
              "Dizziness;Vertigo", "Dysphagia", "Dyspnoea;Death", "Dysuria;Pyrexia", "Encephalitis",  
              "Encephalopathy", "Encephalopathy;Pyrexia", "Epididymitis", "Epilepsy", "Erythema;Dermatitis",  
              "Fall;Malaise", "Fatigue", "Fatigue;Diarrhoea;Nausea;Gastroenteritis;Headache;Malaise;Cryptococcosis;Pyrexia",  
              "Fatigue;Nausea;Malnutrition;Hypophagia", "Giardiasis", "Gout", "Haemarthrosis",  
              "Haemorrhage", "Hepatotoxicity", "Hypercreatinaemia", "Hyperglycaemia", "Hyperkalaemia",  
              "Hyperkalaemia;Hyperphosphataemia", "Hyperphosphataemia", "Hypertension", "Hypotension",  
              "Hypotension;Vomiting;Nausea", "Insomnia;Hypothermia;Pain", "Ischaemia;Infarction",  
              "Neutropenia;Pericarditis", "Neutropenia;Pyrexia;Gastroenteritis",  
              "Oedema;Paraesthesia", "Osteomyelitis;Pain", "Pancreatitis", "Pancreatitis;Pneumonia",  
              "Pancytopenia", "Papilloedema", "Pericarditis", "Pharyngitis",  
              "Pneumonia;Covid-19", "Pneumonia;Fatigue;Covid-19", "Pneumonia;Influenza",  
              "Pneumonia;Pancreatitis", "Pneumonia;Pyrexia;Dyspnoea;Pneumonitis", "Pneumonitis",  
              "Pneumonitis;Bacteraemia;Sepsis", "Pollakiuria", "Pruritus;Rash", "Pyrexia",  
              "Pyrexia;Dysphagia", "Pyrexia;Influenza", "Rash", "Seizure", "Sepsis;Bacteraemia;Pneumonitis",  
              "Sepsis;Cholelithiasis", "Sinusitis", "Splenomegaly;Lymphadenopathy;Infection;Cough;Lymphadenitis;Neutropenia",  
              "Syncope", "Tachyarrhythmia", "Thrombocytopenia", "Thrombocytopenia;Leukopenia", "Vomiting")

freq <- c(18, 10, 7, 4, 4,  
          3, 3, 3, 3, 3,  
          0, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 0, 1, 1, 1,  
          1, 1, 1, 0, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  
          1, 1, 1, 1, 1,  1, 1, 1, 1, 1,
          1, 1, 3, 4)

# Create the Reaction variable as a vector
Reaction <- unlist(mapply(rep, reaction, freq))



# Variable 5: Outcomes
outcomes <- c(
  "Died", "Died;Hospitalized", "Hospitalized", "Hospitalized;Died", "Hospitalized;Disabled", "Non-Serious")
percentages <- c(11.8, 2.96, 75.6, 2.22, 0.74, 6.67) # setting probability same as original
probabilities <- percentages / 100
synthetic$Outcome <- sample(outcomes, size = nrow(synthetic), replace = TRUE, prob = probabilities)

# Variable 6: Active Ingredient
synthetic$activeIngredient <- purrr::map_chr(sample(c("Ibrutinib;Obinutuzumab;Venetoclax",
  "Ibrutinib;Prednisolone;Vincristine;Rituximab;Cyclophosphamide;Venetoclax;Doxorubicin Hydrochloride",
  "Ibrutinib;Rituximab;Prednisolone;Cyclophosphamide;Doxorubicin Hydrochloride;Venetoclax;Vincristine Sulfate",
  "Ibrutinib;Venetoclax",
  "Ibrutinib;Venetoclax;Obinutuzumab",
  "Obinutuzumab;Ibrutinib;Venetoclax",
  "Obinutuzumab;Venetoclax",
  "Obinutuzumab;Venetoclax;Bendamustine",
  "Obinutuzumab;Venetoclax;Ibrutinib",
  "Rituximab;Venetoclax",
  "Ublituximab;Venetoclax;Umbralisib",
  "Venetoclax",
  "Venetoclax;Bendamustine;Obinutuzumab",
  "Venetoclax;Ibrutinib",
  "Venetoclax;Ibrutinib;Obinutuzumab",
  "Venetoclax;Losartan Potassium",
  "Venetoclax;Obinutuzumab",
  "Venetoclax;Obinutuzumab;Ibrutinib",
  "Venetoclax;Rituximab"), obs, replace = TRUE), as.character)

#Variable 7 : Reason
synthetic$Reason <- purrr::map_chr(sample(c("Chronic Lymphocytic Leukaemia"), obs, replace = TRUE), as.character) # all observation had this same value.

```

## Visualizing the Synthetic data
Now we will visualize the synthetic data following the simillar methodologies by the original author.

### Distribution of Patient Age
```{r}
ggplot(synthetic, aes(x = Age)) +
  geom_histogram(binwidth = 2, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Patient Age", x = "Age", y = "Frequency") +
  theme_minimal()
```
The distribution looks somewhat simillar to the original data for Age


### Distribution of Patient Weight
```{r}
ggplot(synthetic, aes(x = Weight)) +
  geom_histogram(binwidth = 5, fill = "cyan", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Patient Weight", x = "Weight (kg)", y = "Frequency") +
  theme_minimal()
```
The distribution looks somewhat simillar to the original distribution of Weight.

### Percentage distribution of Sex
```{r}
sex_distribution <- synthetic %>%
  count(Sex) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

print("Percentage Distribution of Sex:")
print(sex_distribution)
```
The distribution of Sex has been set similar to original distribution because we knew the actual frequencies.

```{r}
ggplot(synthetic, aes(x = Sex, fill = Sex)) +
  geom_bar() +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count") +
  theme_minimal()
```
The similarity in dataset can be verified by the histogram as well.


### Frequency Table for Top 5 Adverse Reactions

I tried multiple times but could not assign the frequencies properly for the values as there we 84 different values. 
```{r}

# Tabulate the variable Reaction and arrange by descending frequency
reaction_counts <- synthetic %>%
  count(Reaction) %>%               # Count occurrences of each reaction
  arrange(desc(n))                  # Arrange by descending frequency

# Print the top 5 most frequent adverse reactions
print("Top 5 Most Frequent Adverse Reactions:")
print(head(reaction_counts, 5))     # Display the top 5 reactions

```

### Percentage distribution of Outcomes
```{r}
outcome_distribution <- synthetic %>%
  count(Outcome) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

print("Percentage Distribution of Outcomes:")
print(outcome_distribution)
```
We are also able to replicate the percentage distribution of the Outcomes as we know the actual frequencies and percentage.

### Bar plot for Outcome frequencies
```{r}
ggplot(synthetic, aes(x = Outcome, fill = Outcome)) +
  geom_bar() +
  labs(title = "Distribution of Outcomes", x = "Outcome", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The similarity in frequencies of outcomes can be verified through the colored histogram as well.